$Z^\prime$ mass limits and the naturalness of supersymmetry

### 2015-03-31 paper/e6ssm-tuning-paper.tex
Added alternative colour palette plots and changed text to be consistent with new colours

\begin{figure}[h]
\begin{center}
\resizebox{!}{6.1cm}{%
\includegraphics{mssm_stop_tuning.png}}
\resizebox{!}{6.1cm}{%
\includegraphics{mssm_higgs_tuning.png}}
\caption{\textit{Left panel:} Scatter plot of fine tuning in the MSSM as a function of
the lightest stop mass, $m_{\tilde{t}_1}$, for the cutoff scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV. 
\textit{Right panel:} Scatter plot of fine tuning in the MSSM as a function of
the lightest Higgs mass, $m_{h_1}$, for the cutoff scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV.}
\label{Fig:mssmstoptuning}
\end{center}
\end{figure}
To illustrate this, in the left panel of Fig.~\ref{Fig:mssmstoptuning}
we show the variation in fine tuning for $M_X=20$ TeV, $50$ TeV, $100$
TeV and $10^{16}$ GeV when we scan over the stop masses and mixing,
with $500 \textrm{ GeV} \leq m_{Q_3},m_{u_3}\leq 10\textrm{ TeV}$ and
$-3810\textrm{ GeV}\leq A_t\leq -20\textrm{ GeV}$.  The remaining
parameters we fix such that at $M_{\rm SUSY}$ they have the values
$\mu=-97.5$ GeV, $B=-84.8$ GeV, $M_1=92.1$ GeV, $M_2=95.9$ GeV,
$M_3=352$ GeV, $A_b=-117.9$ GeV, $A_\tau=-7.8$ GeV, $m_{L_i}=400$ GeV,
$m_{e_i}=204$ GeV, $m_{Q_{1,2}}=438$ GeV, $m_{u_{1,2}}436$ GeV and
$m_{d_i}=438$ GeV ($i=1,2,3$). Here we denote by $M_1$, $M_2$ and $M_3$
the soft gaugino masses for $U(1)$, $SU(2)$ and $SU(3)$, while $A_b$ and
$A_\tau$ are soft trilinear couplings and the $m_\phi$ are soft scalar
masses for the indicated fields. The soft bilinear
coupling $B$ is defined such that at tree level the mass of the CP-odd MSSM Higgs boson
reads $m_A^2=2B\mu/\sin 2\beta$. All off-diagonal couplings and scalar masses
are set to zero, as are the first and second generation Yukawa couplings
and soft trilinears. Although we should stress that making
this choice will lead to a spectrum which is in conflict with the LHC
limits, doing so ensures that fine tuning due to the other parameters
is small, so that we avoid washing out the fine tuning impact of the
stops when the tuning is small\footnote{For models in which the
  spectrum is heavier, when the stop masses are small the fine tuning
  reaches a lower bound imposed by other heavier parameters.} as can
be the case when the stop masses are less than 1 TeV.  Note that the
Higgs mass is also allowed to vary in this scan, as shown in the right
panel of Fig.~\ref{Fig:mssmstoptuning}. This illustrates the tuning
problem which people have been worrying about since the discovery of
the 125 GeV Higgs boson as we see that raising the stop masses is also
pushing up the Higgs mass, meaning that heavier Higgs masses require
more fine tuning.  However for a low value of the UV scale this tuning
is not so severe unless the stops are very heavy, and a 125 GeV Higgs
can be obtained without much tuning in this unrealistic case where we
have minimised other sources of tuning.  On the other hand the tuning
becomes more severe as we increase the cutoff such that for $M_X =
10^{16}$ GeV a lightest stop masses of $1-3$ TeV can result in a fine
tuning of $\approx 100 - 1000$ and the minimum tuning we see
consistent with a $125$ GeV Higgs is $\approx 200$, as shown in
Fig.~\ref{Fig:mssmstoptuning}.

@@

\begin{figure}[h]
\begin{center}
\resizebox{!}{10cm}{%
\includegraphics{all_models_tuning_vs_higgs.png}}
\caption{Scatter plot of fine tuning vs lightest Higgs mass for the
  MSSM (green, bottom band), E$_6$SSM with $M_{Z^\prime} = 2.5$ TeV (blue, middle band) and
  E$_6$SSM with $M_{Z^\prime} = 4.5$ TeV (red, top band). Note that there are points for which
the fine tuning in the MSSM and E$_6$SSM with $M_{Z'}=2.5$ TeV is larger than is visible on this plot
and those below;
however these points are obscured by the overlaid data for the E$_6$SSM with $M_{Z'}=4.5$ TeV, and it is the 
lower bound on the achievable tuning that is of interest here.}
\label{Fig:e6ssmvsmssm}
\end{center}
\end{figure}
In Fig.~\ref{Fig:e6ssmvsmssm} results from the scan are plotted showing
the tuning for each case against the lightest Higgs mass.  As expected
the dependence on the Higgs mass is now quite weak, while the minimum
tuning in the model for the E$_6$SSM is increased by the mass of the
$Z^\prime$ boson.  So in the case of a very low cutoff the tuning
required to get a 125 GeV Higgs is not so large.  However the tuning
from the $Z^\prime$ mass appears already at tree level and is
therefore not suppressed when the cutoff scale is low. In our scan we
find that, for the points satisfying the current limit on the mass of
the $Z^\prime$ boson and having an approximately 125 GeV Higgs, the
minimum fine tuning that can be achieved is
$\Delta_{\textrm{min}}\approx 121$. If run II of the LHC further
pushes up the limit on the $Z^\prime$ mass to be above $4.5$ TeV then
the fine tuning in the model will be greater than at least
$\Delta_{\textrm{min}}\approx 394$ for a Higgs mass between $124.5$
GeV and $125.5$ GeV.

@@

\begin{figure}[h]
\begin{center}
\resizebox{!}{10cm}{%
\includegraphics{mssm_chargino_plot.png}}%%
\caption{Scatter plot of fine tuning vs lightest Higgs mass in the MSSM with $m_{\tilde{\chi}^\pm_1} = 200$ GeV shown in green 
(bottom band) $m_{\tilde{\chi}^\pm_1} = 500$ GeV in blue (middle band) and $m_{\tilde{\chi}^\pm_1} = 700$ GeV in red
(top band).}
\label{Fig:chargino-plot}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\resizebox{!}{6.1cm}{%
\includegraphics{e6inert_tb10_tuning_vs_higgs.png}}
\resizebox{!}{6.1cm}{%
\includegraphics{e6psi_tb10_tuning_vs_higgs.png}}\\
\resizebox{!}{6.1cm}{%
\includegraphics{e6eta_tb10_tuning_vs_higgs.png}}%%
\caption{\textit{Top left panel:} Scatter plot of the fine tuning vs lightest Higgs mass in the $U(1)_I$ model. \textit{Top right panel:} Scatter plot of the fine tuning vs lightest Higgs mass in the $U(1)_\psi$ model. \textit{Bottom panel:} Scatter plot of the fine tuning vs lightest Higgs mass in the $U(1)_\eta$ model. In each plot points with $M_{Z^\prime} = 2.5$ TeV are shown in blue
(bottom band), and points with $M_{Z^\prime} = 4.5$ TeV are shown in red (top band).}
\label{Fig:othere6modelsvsmssm}
\end{center}
\end{figure}

@@

\includegraphics{e6ssm_mssm_cutoff_scan_4_benchmark_points_tuning_vs_cutoff.png} } \\

@@

\includegraphics{mssm_cutoff_scan_tb10_fixed_at_ms_BM1_all_tuning_data.png}} 

@@

\includegraphics{e6ssm_light_zprime_cutoff_scan_tb10_fixed_at_ms_BM1_all_tuning_data.png}} \\

@@

\includegraphics{mssm_cutoff_scan_tb10_fixed_at_ms_BM2_all_tuning_data.png}} 

@@

\includegraphics{e6ssm_light_zprime_cutoff_scan_tb10_fixed_at_ms_BM2_all_tuning_data.png}}%%

@@

For E$_6$SSM BM1 the fine tuning is shown by the solid curve in the top
panel of Fig.~\ref{Fig:BMs-varyMX} and the individual sensitivities
are given in the middle right panel. The tuning actually reduces
initially as the cutoff is is increased from $20$ TeV.  This occurs
because the largest sensitivity is initially $\Delta_\lambda$ (shown
in solid blue in the middle right panel).  This contains some terms
proportional to $M_Z^{\prime \, 2}$, which provide the dominant
contribution to this sensitivity at very low $M_X$.  However as $M_X$
is increased contributions from the soft masses become more important
and these actually start to cancel the large contribution to
$\Delta_\lambda$ coming from $M_{Z^\prime}$ until $\Delta_\lambda$
passes through zero.  At the same time though these large soft masses
also cause other sensitivities to grow, in particular $\Delta_{M_3}$.
The fine tuning rises with $M_X$ once $M_X \gtrsim 10^5-10^6$ GeV, but
remains lower than that of the other points, until $M_X \approx 10^8$
GeV.  Eventually the $\Delta_{M_3}$ sensitivity leads to this point
being the most fine tuned of the four shown in
Fig.~\ref{Fig:BMs-varyMX}.

>>>>

\begin{figure}[h]
\begin{center}
\resizebox{!}{6.1cm}{%
\includegraphics{mssm_stop_tuning_bw_var.png}}
\resizebox{!}{6.1cm}{%
\includegraphics{mssm_higgs_tuning_bw_var.png}}
\caption{\textit{Left panel:} Scatter plot of fine tuning in the MSSM as a function of
the lightest stop mass, $m_{\tilde{t}_1}$, for the cutoff scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV. 
\textit{Right panel:} Scatter plot of fine tuning in the MSSM as a function of
the lightest Higgs mass, $m_{h_1}$, for the cutoff scales (from 
bottom to top) $M_X=20$ TeV, $M_X=50$ TeV, $M_X=100$ TeV and $M_X=10^{16}$ GeV.}
\label{Fig:mssmstoptuning}
\end{center}
\end{figure}
To illustrate this, in the left panel of Fig.~\ref{Fig:mssmstoptuning}
we show the variation in fine tuning for $M_X=20$ TeV, $50$ TeV, $100$
TeV and $10^{16}$ GeV when we scan over the stop masses and mixing,
with $500 \textrm{ GeV} \leq m_{Q_3},m_{u_3}\leq 10\textrm{ TeV}$ and
$-3810\textrm{ GeV}\leq A_t\leq -20\textrm{ GeV}$.  The remaining
parameters we fix such that at $M_{\rm SUSY}$ they have the values
$\mu=-97.5$ GeV, $B=-84.8$ GeV, $M_1=92.1$ GeV, $M_2=95.9$ GeV,
$M_3=352$ GeV, $A_b=-117.9$ GeV, $A_\tau=-7.8$ GeV, $m_{L_i}=400$ GeV,
$m_{e_i}=204$ GeV, $m_{Q_{1,2}}=438$ GeV, $m_{u_{1,2}}=436$ GeV and
$m_{d_i}=438$ GeV ($i=1,2,3$). Here we denote by $M_1$, $M_2$ and $M_3$
the soft gaugino masses for $U(1)$, $SU(2)$ and $SU(3)$, while $A_b$ and
$A_\tau$ are soft trilinear couplings and the $m_\phi$ are soft scalar
masses for the indicated fields. The soft bilinear
coupling $B$ is defined such that at tree level the mass of the CP-odd MSSM Higgs boson
reads $m_A^2=2B\mu/\sin 2\beta$. All off-diagonal couplings and scalar masses
are set to zero, as are the first and second generation Yukawa couplings
and soft trilinears. Although we should stress that making
this choice will lead to a spectrum which is in conflict with the LHC
limits, doing so ensures that fine tuning due to the other parameters
is small, so that we avoid washing out the fine tuning impact of the
stops when the tuning is small\footnote{For models in which the
  spectrum is heavier, when the stop masses are small the fine tuning
  reaches a lower bound imposed by other heavier parameters.} as can
be the case when the stop masses are less than 1 TeV.  Note that the
Higgs mass is also allowed to vary in this scan, as shown in the right
panel of Fig.~\ref{Fig:mssmstoptuning}. This illustrates the tuning
problem which people have been worrying about since the discovery of
the 125 GeV Higgs boson as we see that raising the stop masses is also
pushing up the Higgs mass, meaning that heavier Higgs masses require
more fine tuning.  However for a low value of the UV scale this tuning
is not so severe unless the stops are very heavy, and a 125 GeV Higgs
can be obtained without much tuning in this unrealistic case where we
have minimised other sources of tuning.  On the other hand the tuning
becomes more severe as we increase the cutoff such that for $M_X =
10^{16}$ GeV a lightest stop masses of $1-3$ TeV can result in a fine
tuning of $\approx 100 - 1000$ and the minimum tuning we see
consistent with a $125$ GeV Higgs is $\approx 200$, as shown in
Fig.~\ref{Fig:mssmstoptuning}.

@@

\begin{figure}[h]
\begin{center}
\resizebox{!}{10cm}{%
\includegraphics{all_models_tuning_vs_higgs_bw_var.png}}
\caption{Scatter plot of fine tuning vs lightest Higgs mass for the
  MSSM (light blue, bottom band), E$_6$SSM with $M_{Z^\prime} = 2.5$ TeV (dark blue, middle band) and
  E$_6$SSM with $M_{Z^\prime} = 4.5$ TeV (dark yellow, top band). Note that there are points for which
the fine tuning in the MSSM and E$_6$SSM with $M_{Z'}=2.5$ TeV is larger than is visible on this plot
and those below;
however these points are obscured by the overlaid data for the E$_6$SSM with $M_{Z'}=4.5$ TeV, and it is the 
lower bound on the achievable tuning that is of interest here.}
\label{Fig:e6ssmvsmssm}
\end{center}
\end{figure}
In Fig.~\ref{Fig:e6ssmvsmssm} results from the scan are plotted showing
the tuning for each case against the lightest Higgs mass.  As expected
the dependence on the Higgs mass is now quite weak, while the minimum
tuning in the model for the E$_6$SSM is increased by the mass of the
$Z^\prime$ boson.  So in the case of a very low cutoff the tuning
required to get a 125 GeV Higgs is not so large.  However the tuning
from the $Z^\prime$ mass appears already at tree level and is
therefore not suppressed when the cutoff scale is low. In our scan we
find that, for the points satisfying the current limit on the mass of
the $Z^\prime$ boson and having an approximately 125 GeV Higgs, the
minimum fine tuning that can be achieved is
$\Delta_{\textrm{min}}\approx 121$. If run II of the LHC further
pushes up the limit on the $Z^\prime$ mass to be above $4.5$ TeV then
the fine tuning in the model will be greater than at least
$\Delta_{\textrm{min}}\approx 394$ for a Higgs mass between $124.5$
GeV and $125.5$ GeV.

@@

\begin{figure}[h]
\begin{center}
\resizebox{!}{10cm}{%
\includegraphics{mssm_chargino_plot_bw_var.png}}%%
\caption{Scatter plot of fine tuning vs lightest Higgs mass in the MSSM with $m_{\tilde{\chi}^\pm_1} = 200$ GeV shown in light blue 
(bottom band), $m_{\tilde{\chi}^\pm_1} = 500$ GeV in dark blue (middle band), and $m_{\tilde{\chi}^\pm_1} = 700$ GeV in dark yellow
(top band).}
\label{Fig:chargino-plot}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\resizebox{!}{6.1cm}{%
\includegraphics{e6inert_tb10_tuning_vs_higgs_bw_var.png}}
\resizebox{!}{6.1cm}{%
\includegraphics{e6psi_tb10_tuning_vs_higgs_bw_var.png}}\\
\resizebox{!}{6.1cm}{%
\includegraphics{e6eta_tb10_tuning_vs_higgs_bw_var.png}}%%
\caption{\textit{Top left panel:} Scatter plot of the fine tuning vs lightest Higgs mass in the $U(1)_I$ model. \textit{Top right panel:} Scatter plot of the fine tuning vs lightest Higgs mass in the $U(1)_\psi$ model. \textit{Bottom panel:} Scatter plot of the fine tuning vs lightest Higgs mass in the $U(1)_\eta$ model. In each plot points with $M_{Z^\prime} = 2.5$ TeV are shown in dark blue
(bottom band), and points with $M_{Z^\prime} = 4.5$ TeV are shown in dark yellow (top band).}
\label{Fig:othere6modelsvsmssm}
\end{center}
\end{figure}

@@

\includegraphics{e6ssm_mssm_cutoff_scan_4_benchmark_points_tuning_vs_cutoff_bw_var.png} } \\

@@

\includegraphics{mssm_cutoff_scan_tb10_fixed_at_ms_BM1_all_tuning_data_bw_var.png}} 

@@

\includegraphics{e6ssm_light_zprime_cutoff_scan_tb10_fixed_at_ms_BM1_all_tuning_data_bw_var.png}} \\

@@

\includegraphics{mssm_cutoff_scan_tb10_fixed_at_ms_BM2_all_tuning_data_bw_var.png}} 

@@

\includegraphics{e6ssm_light_zprime_cutoff_scan_tb10_fixed_at_ms_BM2_all_tuning_data_bw_var.png}}%%

@@

For E$_6$SSM BM1 the fine tuning is shown by the solid curve in the top
panel of Fig.~\ref{Fig:BMs-varyMX} and the individual sensitivities
are given in the middle right panel. The tuning actually reduces
initially as the cutoff is is increased from $20$ TeV.  This occurs
because the largest sensitivity is initially $\Delta_\lambda$ (shown
in solid light blue in the middle right panel).  This contains some terms
proportional to $M_Z^{\prime \, 2}$, which provide the dominant
contribution to this sensitivity at very low $M_X$.  However as $M_X$
is increased contributions from the soft masses become more important
and these actually start to cancel the large contribution to
$\Delta_\lambda$ coming from $M_{Z^\prime}$ until $\Delta_\lambda$
passes through zero.  At the same time though these large soft masses
also cause other sensitivities to grow, in particular $\Delta_{M_3}$.
The fine tuning rises with $M_X$ once $M_X \gtrsim 10^5-10^6$ GeV, but
remains lower than that of the other points, until $M_X \approx 10^8$
GeV.  Eventually the $\Delta_{M_3}$ sensitivity leads to this point
being the most fine tuned of the four shown in
Fig.~\ref{Fig:BMs-varyMX}.
